### 2019-01-11 write-up/dissertation/background.tex
Add title

The syntax is intended to resemble OCaml's, apart from the spurious
\ltfla{`!'}s found here and there (they are annotations to show that we can
use a value more than once). In Figure~\ref{fig:ltfla_sumarray}, we see
\ltfla{row} has type (\ltfla{'x arr}). More detailed explanations of
\emph{what and why} will be given in Chapter~\ref{chap:impl}, but for now, it
is enough to know it means we can only \emph{read} from \ltfla{row}, and not
write to it. If we did try to write to or free \ltfla{row}, we would get a
helpful error message, as shown in Figure~\ref{fig:ltfla_univ}.

@@

Now suppose we are trying to square a matrix, using a `dgemm' like BLAS
routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
assume `simple\_dgemm' has type \ltfla{'x . 'x mat --o 'y .  'y mat --o z mat
--o ('x mat * 'y mat) * z mat}.} in Figures~\ref{fig:ltfla_square}
and~\ref{fig:ltfla_free}) which takes two read-only matrices and a third matrix
it can write to and performs $C := AB + C$. How would we use such a routine to
square a matrix, assuming $C=0$ initially?  Surely this would break linearity,
since $A$ would have to be passed in to the function twice, and we can only 
use non-!-annotated variables once?

To solve this, we use a special primitive called \ltfla{share} to produce
(more) read-only \emph{aliases} of any matrix. We then pass these into the
function, and it works as expected. Once the squared matrix has been obtained,
we may not want the original anymore, and thus decide to free it. Before we do
so, we must first \ltfla{unshare} any read-only aliases that exist, to get back
a single, read-write handle.

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let a = unshare _ a1 a2 in
let () = free a in c
    \end{minted}
    \caption{Squaring a matrix in LT4LA (assuming \ltfla{c} is initially
        0).}\label{fig:ltfla_square}
\end{figure}

If we tried to free one of the read-only aliases before or instead of
\ltfla{unshare}-ing them, then we would get the error shown in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
Appendix~\ref{chap:primitives}) are set up so that you can only free something
when you have read-write access to it, guaranteed by linearity to be the only
name in scope with this capability (aliases can only be read-only). Conversely,
if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
used} error.

Another way in which `simple\_dgemm' may be misused is by instantiating its
informal description of $C := AB + C$ with $B = C = A$ and so mistakenly
concluding that it computes $A := A^2 + A$ in-place. However, the type of
\ltfla{share} prevents this as well -- \ltfla{let (a11, a12) = share _ a1 in
simple_dgemm _ a11 _ a12 a2} would result in an error similar to the one shown
in Figure~\ref{fig:ltfla_free}.

>>>>

The syntax is intended to resemble OCaml's, apart from the spurious
\highl{`!'}s found here and there (they are annotations to show that we can
use a value more than once). In Figure~\ref{fig:ltfla_sumarray}, we see
\highl{row} has type (\highl{'x arr}). More detailed explanations of
\emph{what and why} will be given in Chapter~\ref{chap:impl}, but for now, it
is enough to know it means we can only \emph{read} from \highl{row}, and not
write to it. If we did try to write to or free \highl{row}, we would get a
helpful error message, as shown in Figure~\ref{fig:ltfla_univ}.

@@

Now suppose we are trying to square a matrix, using a `dgemm' like BLAS
routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
assume `simple\_dgemm' has type \highl{'x . 'x mat --o 'y .  'y mat --o z mat
--o ('x mat * 'y mat) * z mat}.} in Figures~\ref{fig:ltfla_square}
and~\ref{fig:ltfla_free}) which takes two read-only matrices and a third matrix
it can write to and performs $C := AB + C$. How would we use such a routine to
square a matrix, assuming $C=0$ initially?  Surely this would break linearity,
since $A$ would have to be passed in to the function twice, and we can only 
use non-!-annotated variables once?

To solve this, we use a special primitive called \highl{share} to produce
(more) read-only \emph{aliases} of any matrix. We then pass these into the
function, and it works as expected. Once the squared matrix has been obtained,
we may not want the original anymore, and thus decide to free it. Before we do
so, we must first \highl{unshare} any read-only aliases that exist, to get back
a single, read-write handle.

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let a = unshare _ a1 a2 in
let () = free a in c
    \end{minted}
    \caption{Squaring a matrix in LT4LA (assuming \highl{c} is initially
        0).}\label{fig:ltfla_square}
\end{figure}

If we tried to free one of the read-only aliases before or instead of
\highl{unshare}-ing them, then we would get the error shown in
Figure~\ref{fig:ltfla_free}. Briefly, a \highl{z arr} is a read-write array,
everything else (\highl{'x arr} or \highl{_ s arr}) is read-only. The types of
\highl{free}, \highl{share} and \highl{unshare} (see
Appendix~\ref{chap:primitives}) are set up so that you can only free something
when you have read-write access to it, guaranteed by linearity to be the only
name in scope with this capability (aliases can only be read-only). Conversely,
if we \emph{forgot} to \highl{free a}, we would also get \texttt{Variable a not
used} error.

Another way in which `simple\_dgemm' may be misused is by instantiating its
informal description of $C := AB + C$ with $B = C = A$ and so mistakenly
concluding that it computes $A := A^2 + A$ in-place. However, the type of
\highl{share} prevents this as well -- \highl{let (a11, a12) = share _ a1 in
simple_dgemm _ a11 _ a12 a2} would result in an error similar to the one shown
in Figure~\ref{fig:ltfla_free}.

### 2019-01-11 write-up/dissertation/evaluation.tex
Add Appendix and more figures
Add title

I measured execution time, in micro-seconds, against an exponentially (powers
of 5) increasing scaling factor for matrix size parameters $n=5$ and $k=3$.
For small scaling factors (1, 5, 25), I used the Core\_bench micro-benchmarking
library, for larger factors (125 and greater), I used the \texttt{getrusage}
system call (called \ltfla{Unix.times} in OCaml), sandwiched between calls to
\ltfla{Gc.full_major} to minimise the effects of garbage-collection.
Core\_bench performs a linear regression (here, time against batch-size) so
includes a 95\% confidence-interval with $R^2$. Larger scales have
errors reported to $\pm \sigma$ (one standard-deviation).

@@

\begin{figure}[t]
    \centering
    \input{timings.tex}
    \caption{Comparison of execution times (error bars are present but quite
        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
        with the Core\_bench library. Larger ones used Unix's
        \texttt{getrusage} functionality, sandwiched between calls to
        \ltfla{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
\end{figure}

@@

Having access to primitives which allow a programmer to re-use memory means
that memory usage for temporaries in LT4LA is on par with that of CBLAS. One
caveat is that the \ltfla{freeM} primitive is a no-op in LT4LA, so deallocation
still relies on OCaml's garbage-collector.

@@

\begin{figure}[tp]
    \centering
    \input{timings_all.tex}
    \caption{Comparison of execution times (error bars are present but quite
        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
        with the Core\_bench library. Larger ones used Unix's
        \texttt{getrusage} functionality, sandwiched between calls to
        \ltfla{Gc.full_major} for the OCaml implementations.}\label{fig:timings_all}
\end{figure}

@@

The trend here is that for $n=625$ and $n=3125$ (sizes for which
\ltfla{Gc.full_major} was called before each measurement), mean execution times
start to split into two groups: Chol and Owl as one group (similar,
\emph{faster} times) and LT4LA and CBLAS as another (similar, \emph{slower}
times). \emph{Although the goals of this thesis have been met}, this is
unexpected behaviour, and attempting to understand \emph{why} the Chol/Owl
implementations tended to be faster for anything except the smallest of
matrices, baffled several people I consulted, including experts at OCamlLabs.

@@

The original implementation used 6 temporaries, one of which was immediately
spotted as never being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional-capabilities. Lastly, it contained another unnecessary
temporary, however one that was not obvious without linear types. To spot it, I
first performed live-range splitting (checked by linearity) by hoisting calls
to \ltfla{freeM} and then annotated the freed matrices with their dimensions.
After doing so and spotting two disjoint live-ranges of the same size, I
replaced a call to \ltfla{freeM} followed by allocating call to \ltfla{copy}
with one, in-place call to \ltfla{copyM_to}. I believe the ability to boldly
refactor code which manages memory is good evidence of the usefulness of
linearity as a tool for programming.

>>>>

I measured execution time, in micro-seconds, against an exponentially (powers
of 5) increasing scaling factor for matrix size parameters $n=5$ and $k=3$.
For small scaling factors (1, 5, 25), I used the Core\_bench micro-benchmarking
library, for larger factors (125 and greater), I used the \texttt{getrusage}
system call (called \highl{Unix.times} in OCaml), sandwiched between calls to
\highl{Gc.full_major} to minimise the effects of garbage-collection.
Core\_bench performs a linear regression (here, time against batch-size) so
includes a 95\% confidence-interval with $R^2$. Larger scales have
errors reported to $\pm \sigma$ (one standard-deviation).

@@

\begin{figure}[t]
    \centering
    \input{timings.tex}
    \caption{Comparison of execution times (error bars are present but quite
        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
        with the Core\_bench library. Larger ones used Unix's
        \texttt{getrusage} functionality, sandwiched between calls to
        \highl{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
\end{figure}

@@

Having access to primitives which allow a programmer to re-use memory means
that memory usage for temporaries in LT4LA is on par with that of CBLAS. One
caveat is that the \highl{freeM} primitive is a no-op in LT4LA, so deallocation
still relies on OCaml's garbage-collector.

@@

\begin{figure}[tp]
    \centering
    \input{timings_all.tex}
    \caption{Comparison of execution times (error bars are present but quite
        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
        with the Core\_bench library. Larger ones used Unix's
        \texttt{getrusage} functionality, sandwiched between calls to
        \highl{Gc.full_major} for the OCaml implementations.}\label{fig:timings_all}
\end{figure}

@@

The trend here is that for $n=625$ and $n=3125$ (sizes for which
\highl{Gc.full_major} was called before each measurement), mean execution times
start to split into two groups: Chol and Owl as one group (similar,
\emph{faster} times) and LT4LA and CBLAS as another (similar, \emph{slower}
times). \emph{Although the goals of this thesis have been met}, this is
unexpected behaviour, and attempting to understand \emph{why} the Chol/Owl
implementations tended to be faster for anything except the smallest of
matrices, baffled several people I consulted, including experts at OCamlLabs.

@@

The original implementation used 6 temporaries, one of which was immediately
spotted as never being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional-capabilities. Lastly, it contained another unnecessary
temporary, however one that was not obvious without linear types. To spot it, I
first performed live-range splitting (checked by linearity) by hoisting calls
to \highl{freeM} and then annotated the freed matrices with their dimensions.
After doing so and spotting two disjoint live-ranges of the same size, I
replaced a call to \highl{freeM} followed by allocating call to \highl{copy}
with one, in-place call to \highl{copyM_to}. I believe the ability to boldly
refactor code which manages memory is good evidence of the usefulness of
linearity as a tool for programming.

### 2019-01-11 write-up/dissertation/implementation.tex
Organise write-ups
Add more evaluation examples
Add title

To make LT4LA a usable DSL, I needed a way of allowing values to either never
be used or be used more than once. For this, I added the \mbox{!-constructor}
at the type-level of the DSL and its corresponding introduction (the
\ltfla{Many}-constructor) and elimination (the \ltfla{let Many <id> = .. in ..
} destructor) forms at the term-level of the DSL.  The idea behind the
\mbox{!-constructor} is that a value of that type uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a \mbox{!-type}. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to emulate
intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to \ltfla{let Many x =
<exp> in let Many x = Many (Many x) in <body>} (similarly for function argument
bindings). The reader can check (using the rules in
Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.

However, just that desugaring alone is not enough to prevent a user from taking
an array or matrix and moving it into the intuitionistic environments. Why?
There are certain situations in which we \emph{should not} use the \ltfla{Many}
constructor.  Consider the following code: \ltfla{let Many x = Many (array 5)
in <body>}; the expression \ltfla{array 5} uses no linearly-typed variables
from the environment. Although we could just reject types of the form
\ltfla{!(_ arr)} to fix this simple example, what about pairs \ltfla{let Many
xy = Many (3, array 5) in <body>}?  Ad-hoc pattern matching on the type cannot
account for all possible situations. With the last case, we can use \ltfla{xy}
as many times as we would like, destruct the pair to get the second component
and thus create \emph{distinct} read-write aliases to the same array.  Alas,
now arrays can be used intuitionistically and all the benefits of linearity are
lost.  Or are they?

@@

Not quite, but to understand how we can fix this problem, we need to question
an assumption left implicit up until this point: what does \ltfla{Many} even
mean in the DSL? How is it translated to OCaml? What is the OCaml runtime
behaviour of the DSL's \ltfla{Many}-constructor? One option is to go down the
C++ route and make \ltfla{Many} behave like a \ltfla{shared_ptr} and act as a
runtime reference-count for arrays. I chose to not go for this option because
it went against the \emph{explicitness} and \emph{predictability} that C and
Fortran have. It would make analysing when and what is allocated and freed more
like the higher-level languages I was trying to move away from.

My aim is to show linear types are simple to understand and apply, enough so
that they can be grafted (in a limited way) on to existing languages as a
\emph{library}. In that spirit, the simplest thing that the DSL's
\ltfla{Many}-constructor can mean in the OCaml runtime is \emph{nothing}.
LT4LA's !/\ltfla{Many} constructors are translated into OCaml
\ltfla{bang}/\ltfla{Many} constructors declared as \ltfla{type 'a bang = Many
of 'a [@@unboxed]}.  The \emph{unboxed} annotation means that the type and its
constructor only exist for the purpose of type checking in OCaml; \emph{the
OCaml runtime representation of values of type \ltfla{'a} and \ltfla{'a bang}
is exactly the same}.

With this understanding, our problem is that arrays and matrices are unlike
other values such as integers and elements because (under the OCaml hood)
calling a function with an array argument copies a \emph{pointer} to the array
rather than the array itself. So, we can start making a distinction,
\emph{defining} elements, integers, booleans, intuitionistic variables, units
and lambda-expressions that capture no linear variables as \emph{values} (since
they cannot break referential-transparency) and anything else (arrays,
matrices, expressions which can be reduced, such as function application and
if-expressions) as not being `values'. If this sounds familiar, it is because
this is the same \emph{value-restriction} `trick' from the world of polymorphic
types applied to linearity instead. We then have the rule that in LT4LA, we can
only use \ltfla{Many} on expressions that are defined to be \emph{values} and
\emph{use no linear variables}.

@@

Array and matrix types are parameterised by
\emph{fractional-capabilities}~\cite{boyland}.  A fraction of 1 ($2^0$)
represents complete ownership of a value. Creating an array gives you ownership
of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents
`0').  Once you have ownership of an array, you can free it: \ltfla{free : z
arr --o unit}.  Importantly, because a linear value may only be used once, the
array just freed is \emph{out of scope} for following expressions,
\textbf{preventing use-after-free}.  Ownership also enables you to write to the
array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i]
:= j} is just sugar for \ltfla{set w i j}). Here, linearity prevents accessing
aliases which represented the array \emph{before} the mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
means a program can \emph{appear} to use a variable multiple times, important
for keeping LT4LA usable and readable. The underscore is how a programmer tells
the compiler to automatically \emph{infer} the correct fractional-capability
based on the other arguments passed to the function. In conjunction with the
requirement that function declarations need type-annotations for their
arguments, \textbf{this allows fractional-capabilities to be \emph{correctly
inferred} in \emph{any} LT4LA expression}.

Fractions exist to provide both safe aliasing and read/write permissions, via
the primitives \ltfla{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former,
the two arrays returned (which are just aliases of the given array) can now
only be read from and not written to. If you want to write to this array, you
must use the latter to combine other read-only aliases until you are left with
only one value of type \ltfla{z arr}, guaranteeing no other aliases exists.

@@

A non-recursive function may be used more than once if it does not refer to any
linear variables from the surrounding scope. So, we can desugar something like
\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
let Many f =  Many (fun y : !int -> x + y) in  <body>}.  Recursion is slightly
more complicated: we can desugar the factorial function
(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int, if
(*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many}, must also
not use any linear variables from its surrounding scope, because a
recursive function may evaluate its body multiple times.

@@

\begin{figure}[tp]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA. In contrast with the Fortran
        (Figure~\ref{fig:fortran_kalman}) or C (Figure~\ref{fig:cblas_kalman})
        implementations, this one is readable (equational and algebraic),
        using easy to reason about, value-semantic expressions. In addition to
        that, its \emph{intensional} behaviour and assumptions (regarding
        aliasing, read/write permissions and memory allocation, re-use and
        de-allocation) are also explicitly and precisely described and
        \emph{automatically checkable}.}\label{fig:ltfla_kalman}
\end{figure}

@@

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let x <- [| y |] in ..              ==> let (y,x) = copyM_to _ y x in ..
let x <- new [| y |] in ..          ==> let (y,x) = copyM _ y in ..
let x <- [| a*b + c |] in ..        ==> let x <- [| 1.*a*b + 1.*c |] in ..
let x <- [| i*a*b + j*c |] in ..    ==> let ((a,b), c) = (* BLAS *) in ..
let x <- new (m, n) [| a*b |] in .. ==> let c = matrix m n in
                                        let x <- [| 1.*a*b + 0.*c |] in ..
    \end{minted}
    \caption{Syntactic translations of matrix expressions to linearly-typed
        matrix functions. Further annotations on the matrix variables
        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine is called and
        what parameters it is passed.}\label{fig:mat_patterns}

@@

Code generation is a straightforward mapping from core LT4LA constructs to
OCaml constructs, with the addition of \ltfla{Many} constructors to wrap
integer, element and boolean literals.

@@

    \item \ltfla{let xy, z = <exp> in let x,y = xy in <body>} with
        \ltfla{let (x,y), z = <exp> in <body>},

    \item \ltfla{let Many x = x in let Many x = Many (Many x) in <body>} with
        \ltfla{<body>},

    \item \ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>} with
        \ltfla{let x = <exp> in <body>},

    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body>},

    \item \ltfla{let Many f = fix (f, x, .., <exp>, ..) in <body>} with
        \ltfla{let rec f x = <exp> in <body>}.

@@

I explained how a few core features -- linearity, the \ltfla{Many}
constructor, value-restriction, fractional-capabilities with inference,
if-expressions and recursive functions -- are enough to \emph{statically
capture and automatically check} aliasing, read/write permissions, memory
allocation, re-use and deallocation of non-trivial linear algebra programs. I
also demonstrated that simple pattern-matching and desugaring provides the
potential for a \textbf{new, \emph{type-directed}} approach to matrix expression
compilation. Lastly I have shown that it is possible to use these features with
\emph{existing} languages and frameworks.\footnote{As mentioned in the previous
chapter, if the host language supports \emph{syntax-extensions}, like PPX
for OCaml, it is possible to construct LT4LA expressions \emph{from within} the
host language.}

>>>>

To make LT4LA a usable DSL, I needed a way of allowing values to either never
be used or be used more than once. For this, I added the \mbox{!-constructor}
at the type-level of the DSL and its corresponding introduction (the
\highl{Many}-constructor) and elimination (the \highl{let Many <id> = .. in ..
} destructor) forms at the term-level of the DSL.  The idea behind the
\mbox{!-constructor} is that a value of that type uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a \mbox{!-type}. This
includes integers, elements and booleans. So \highl{3 : !int} and \highl{3. +.
4. : !elt}. However, all bindings are still linear by default, so to emulate
intuitionism, I desugar \highl{let !x = <exp> in <body>} to \highl{let Many x =
<exp> in let Many x = Many (Many x) in <body>} (similarly for function argument
bindings). The reader can check (using the rules in
Appendix~\ref{chap:ott_spec}) that this has the effect of moving \highl{x : !t}
from the linear to the intuitionistic environments, only if \highl{<exp> : !t}.

However, just that desugaring alone is not enough to prevent a user from taking
an array or matrix and moving it into the intuitionistic environments. Why?
There are certain situations in which we \emph{should not} use the \highl{Many}
constructor.  Consider the following code: \highl{let Many x = Many (array 5)
in <body>}; the expression \highl{array 5} uses no linearly-typed variables
from the environment. Although we could just reject types of the form
\highl{!(_ arr)} to fix this simple example, what about pairs \highl{let Many
xy = Many (3, array 5) in <body>}?  Ad-hoc pattern matching on the type cannot
account for all possible situations. With the last case, we can use \highl{xy}
as many times as we would like, destruct the pair to get the second component
and thus create \emph{distinct} read-write aliases to the same array.  Alas,
now arrays can be used intuitionistically and all the benefits of linearity are
lost.  Or are they?

@@

Not quite, but to understand how we can fix this problem, we need to question
an assumption left implicit up until this point: what does \highl{Many} even
mean in the DSL? How is it translated to OCaml? What is the OCaml runtime
behaviour of the DSL's \highl{Many}-constructor? One option is to go down the
C++ route and make \highl{Many} behave like a \highl{shared_ptr} and act as a
runtime reference-count for arrays. I chose to not go for this option because
it went against the \emph{explicitness} and \emph{predictability} that C and
Fortran have. It would make analysing when and what is allocated and freed more
like the higher-level languages I was trying to move away from.

My aim is to show linear types are simple to understand and apply, enough so
that they can be grafted (in a limited way) on to existing languages as a
\emph{library}. In that spirit, the simplest thing that the DSL's
\highl{Many}-constructor can mean in the OCaml runtime is \emph{nothing}.
LT4LA's !/\highl{Many} constructors are translated into OCaml
\highl{bang}/\highl{Many} constructors declared as \highl{type 'a bang = Many
of 'a [@@unboxed]}.  The \emph{unboxed} annotation means that the type and its
constructor only exist for the purpose of type checking in OCaml; \emph{the
OCaml runtime representation of values of type \highl{'a} and \highl{'a bang}
is exactly the same}.

With this understanding, our problem is that arrays and matrices are unlike
other values such as integers and elements because (under the OCaml hood)
calling a function with an array argument copies a \emph{pointer} to the array
rather than the array itself. So, we can start making a distinction,
\emph{defining} elements, integers, booleans, intuitionistic variables, units
and lambda-expressions that capture no linear variables as \emph{values} (since
they cannot break referential-transparency) and anything else (arrays,
matrices, expressions which can be reduced, such as function application and
if-expressions) as not being `values'. If this sounds familiar, it is because
this is the same \emph{value-restriction} `trick' from the world of polymorphic
types applied to linearity instead. We then have the rule that in LT4LA, we can
only use \highl{Many} on expressions that are defined to be \emph{values} and
\emph{use no linear variables}.

@@

Array and matrix types are parameterised by
\emph{fractional-capabilities}~\cite{boyland}.  A fraction of 1 ($2^0$)
represents complete ownership of a value. Creating an array gives you ownership
of it; the function \highl{array : !int --o z arr} (where \highl{z} represents
`0').  Once you have ownership of an array, you can free it: \highl{free : z
arr --o unit}.  Importantly, because a linear value may only be used once, the
array just freed is \emph{out of scope} for following expressions,
\textbf{preventing use-after-free}.  Ownership also enables you to write to the
array: \highl{set : z arr --o !int --o !elt --o z arr} (the syntax \highl{w[i]
:= j} is just sugar for \highl{set w i j}). Here, linearity prevents accessing
aliases which represented the array \emph{before} the mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \highl{'x}
represents a natural number (either a zero \highl{z}, variable \highl{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\highl{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \highl{let !v
<- w[i]} is just sugar for \highl{let (w, !v) = get _ w i}). In general, a
left-arrow \highl{<-} signifies transparent rebinding with returned values: it
means a program can \emph{appear} to use a variable multiple times, important
for keeping LT4LA usable and readable. The underscore is how a programmer tells
the compiler to automatically \emph{infer} the correct fractional-capability
based on the other arguments passed to the function. In conjunction with the
requirement that function declarations need type-annotations for their
arguments, \textbf{this allows fractional-capabilities to be \emph{correctly
inferred} in \emph{any} LT4LA expression}.

Fractions exist to provide both safe aliasing and read/write permissions, via
the primitives \highl{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
\highl{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former,
the two arrays returned (which are just aliases of the given array) can now
only be read from and not written to. If you want to write to this array, you
must use the latter to combine other read-only aliases until you are left with
only one value of type \highl{z arr}, guaranteeing no other aliases exists.

@@

A non-recursive function may be used more than once if it does not refer to any
linear variables from the surrounding scope. So, we can desugar something like
\highl{let x = 3 in let !f (y : !int) = x + y in <body>} to \highl{let x = 3 in
let Many f =  Many (fun y : !int -> x + y) in  <body>}.  Recursion is slightly
more complicated: we can desugar the factorial function
(Figure~\ref{fig:ltfla_factorial}) to \highl{let Many f = fix (f, x : !int, if
(*..*) : !int) in <body> }. However, \highl{fix}, like \highl{Many}, must also
not use any linear variables from its surrounding scope, because a
recursive function may evaluate its body multiple times.

@@

\begin{figure}[tp]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/kalman.lt}
    \caption{Kalman filter in LT4LA. In contrast with the Fortran
        (Figure~\ref{fig:fortran_kalman}) or C (Figure~\ref{fig:cblas_kalman})
        implementations, this one is readable (equational and algebraic),
        using easy to reason about, value-semantic expressions. In addition to
        that, its \emph{intensional} behaviour and assumptions (regarding
        aliasing, read/write permissions and memory allocation, re-use and
        de-allocation) are also explicitly and precisely described and
        \emph{automatically checkable}.}\label{fig:ltfla_kalman}
\end{figure}

@@

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let x <- [| y |] in ..              ==> let (y,x) = copyM_to _ y x in ..
let x <- new [| y |] in ..          ==> let (y,x) = copyM _ y in ..
let x <- [| a*b + c |] in ..        ==> let x <- [| 1.*a*b + 1.*c |] in ..
let x <- [| i*a*b + j*c |] in ..    ==> let ((a,b), c) = (* BLAS *) in ..
let x <- new (m, n) [| a*b |] in .. ==> let c = matrix m n in
                                        let x <- [| 1.*a*b + 0.*c |] in ..
    \end{minted}
    \caption{Syntactic translations of matrix expressions to linearly-typed
        matrix functions. Further annotations on the matrix variables
        (\highl{sym} or \highl{^T}) determine which BLAS routine is called and
        what parameters it is passed.}\label{fig:mat_patterns}

@@

Code generation is a straightforward mapping from core LT4LA constructs to
OCaml constructs, with the addition of \highl{Many} constructors to wrap
integer, element and boolean literals.

@@

    \item \highl{let xy, z = <exp> in let x,y = xy in <body>} with
        \highl{let (x,y), z = <exp> in <body>},

    \item \highl{let Many x = x in let Many x = Many (Many x) in <body>} with
        \highl{<body>},

    \item \highl{let Many x = <exp> in let Many x = Many (Many x) in <body>} with
        \highl{let x = <exp> in <body>},

    \item \highl{let Many x = Many <exp> in <body>} with \highl{let x = <exp> in <body>},

    \item \highl{let Many f = fix (f, x, .., <exp>, ..) in <body>} with
        \highl{let rec f x = <exp> in <body>}.

@@

I explained how a few core features -- linearity, the \highl{Many}
constructor, value-restriction, fractional-capabilities with inference,
if-expressions and recursive functions -- are enough to \emph{statically
capture and automatically check} aliasing, read/write permissions, memory
allocation, re-use and deallocation of non-trivial linear algebra programs. I
also demonstrated that simple pattern-matching and desugaring provides the
potential for a \textbf{new, \emph{type-directed}} approach to matrix expression
compilation. Lastly I have shown that it is possible to use these features with
\emph{existing} languages and frameworks.\footnote{As mentioned in the previous
chapter, if the host language supports \emph{syntax-extensions}, like PPX
for OCaml, it is possible to construct LT4LA expressions \emph{from within} the
host language.}
